---
title: "Which Demographic, Financial, and Repayment-Related Factors Are Significantly Associated with Consumer Credit Default?"
author: "Haris Jamil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Introduction

Consumer credit default is a major concern for financial institutions, as it directly
affects profitability and risk exposure. This project investigates **which demographic,
financial, and repayment-related factors are significantly associated with consumer
credit default**, using the *Default of Credit Card Clients* dataset (30,000 clients,
Taiwan).

The analysis is organized into three parts:

1. **Demographic factors** — Sex, Education, Marriage, Age
2. **Financial factors** — Credit limit, Bill amounts
3. **Repayment-related factors** — Past repayment status, Payment amounts

For each group, appropriate hypothesis tests are used, followed by a combined
**logistic regression model** to identify which factors remain statistically
significant when considered together.

---

# 2. Load Libraries and Data

```{r}
library(dplyr)
library(ggplot2)
library(car)      # for VIF check
library(broom)     # for tidy regression output

# The first row of the raw file is a redundant header (X1, X2, ...), so skip it
df <- read.csv("Copy_of_default_of_credit_card_clients.csv", skip = 1)

# Rename target column for convenience
names(df)[names(df) == "default.payment.next.month"] <- "DEFAULT"

str(df)
```

---

# 3. Data Cleaning

```{r}
# Check for missing values
colSums(is.na(df))

# Recode categorical variables as factors with meaningful labels
df$SEX <- factor(df$SEX, levels = c(1, 2), labels = c("Male", "Female"))

# EDUCATION: 1=Graduate, 2=University, 3=High School, 4=Others
# Values 0, 5, 6 are undocumented/unknown -> grouped into "Others"
df$EDUCATION <- ifelse(df$EDUCATION %in% c(0, 5, 6), 4, df$EDUCATION)
df$EDUCATION <- factor(df$EDUCATION, levels = c(1, 2, 3, 4),
                        labels = c("Graduate", "University", "High School", "Others"))

# MARRIAGE: 1=Married, 2=Single, 3=Others (0 is undocumented -> grouped into Others)
df$MARRIAGE <- ifelse(df$MARRIAGE == 0, 3, df$MARRIAGE)
df$MARRIAGE <- factor(df$MARRIAGE, levels = c(1, 2, 3),
                       labels = c("Married", "Single", "Others"))

df$DEFAULT_LABEL <- factor(df$DEFAULT, levels = c(0, 1), labels = c("No Default", "Default"))

summary(df[, c("SEX", "EDUCATION", "MARRIAGE", "AGE", "LIMIT_BAL", "DEFAULT_LABEL")])
```

---

# 4. Exploratory Data Analysis

### 4.1 Overall Default Rate

```{r}
default_rate <- prop.table(table(df$DEFAULT_LABEL)) * 100
default_rate
```

```{r}
ggplot(df, aes(x = DEFAULT_LABEL, fill = DEFAULT_LABEL)) +
  geom_bar() +
  labs(title = "Distribution of Credit Card Default", x = "", y = "Number of Clients") +
  theme_minimal() +
  theme(legend.position = "none")
```

### 4.2 Credit Limit by Default Status

```{r}
ggplot(df, aes(x = DEFAULT_LABEL, y = LIMIT_BAL, fill = DEFAULT_LABEL)) +
  geom_boxplot() +
  labs(title = "Credit Limit by Default Status", x = "", y = "Credit Limit (NT$)") +
  theme_minimal() +
  theme(legend.position = "none")
```

### 4.3 Default Rate by Education and Marriage

```{r}
ggplot(df, aes(x = EDUCATION, fill = DEFAULT_LABEL)) +
  geom_bar(position = "fill") +
  labs(title = "Default Rate by Education Level", x = "Education", y = "Proportion") +
  theme_minimal()

ggplot(df, aes(x = MARRIAGE, fill = DEFAULT_LABEL)) +
  geom_bar(position = "fill") +
  labs(title = "Default Rate by Marital Status", x = "Marital Status", y = "Proportion") +
  theme_minimal()
```

---

# 5. Hypothesis Testing

## 5.1 Demographic Factors

**H0:** Default is independent of Sex / Education / Marriage.
**H1:** Default is associated with Sex / Education / Marriage.

```{r}
chisq.test(table(df$SEX, df$DEFAULT_LABEL))
chisq.test(table(df$EDUCATION, df$DEFAULT_LABEL))
chisq.test(table(df$MARRIAGE, df$DEFAULT_LABEL))
```

**Age** (continuous) — compare mean age between defaulters and non-defaulters:

**H0:** Mean age is equal for defaulters and non-defaulters.
**H1:** Mean age differs between the two groups.

```{r}
t.test(AGE ~ DEFAULT_LABEL, data = df)
```

## 5.2 Financial Factors

**Credit Limit (LIMIT_BAL)** — compare mean credit limit between groups:

**H0:** Mean credit limit is equal for defaulters and non-defaulters.
**H1:** Mean credit limit differs between the two groups.

```{r}
t.test(LIMIT_BAL ~ DEFAULT_LABEL, data = df)
```

**Bill Amount (BILL_AMT1, most recent month):**

```{r}
t.test(BILL_AMT1 ~ DEFAULT_LABEL, data = df)
```

## 5.3 Repayment-Related Factors

**Most recent repayment status (PAY_0)** — compare mean repayment delay status:

**H0:** Mean repayment status (PAY_0) is equal for defaulters and non-defaulters.
**H1:** Mean repayment status differs between the two groups.

```{r}
t.test(PAY_0 ~ DEFAULT_LABEL, data = df)
```

**Most recent payment amount (PAY_AMT1):**

```{r}
t.test(PAY_AMT1 ~ DEFAULT_LABEL, data = df)
```

---

# 6. Logistic Regression: Combined Model

To assess which factors remain significant **when controlling for all others**, a
multiple logistic regression model is fitted with `DEFAULT` as the outcome variable.

```{r}
model <- glm(DEFAULT ~ LIMIT_BAL + SEX + EDUCATION + MARRIAGE + AGE +
               PAY_0 + PAY_2 + PAY_3 +
               BILL_AMT1 + PAY_AMT1,
             data = df, family = binomial)

summary(model)
```

### 6.1 Odds Ratios (Easier to Interpret)

```{r}
tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
  arrange(p.value)
```

An odds ratio **> 1** means the factor increases the odds of default; **< 1** means it
decreases the odds. Factors with `p.value < 0.05` are statistically significant
predictors of default.

### 6.2 Multicollinearity Check

```{r}
vif(model)
```

---

# 7. Key Findings

```{r, echo=FALSE}
sig_vars <- tidy(model) %>% filter(p.value < 0.05) %>% pull(term)
cat("Statistically significant predictors (p < 0.05):\n")
print(sig_vars)
```

Based on the hypothesis tests and the logistic regression model:

- **Repayment history (PAY_0, PAY_2, PAY_3)** is typically the **strongest** predictor
  of default — clients with a history of delayed payments are significantly more
  likely to default.
- **Credit limit (LIMIT_BAL)** is inversely associated with default — clients with
  lower credit limits show higher default rates.
- **Demographic factors** (Education, Marriage, Age) show a statistically significant
  but comparatively **weaker** association with default than repayment behavior.
- **Bill and payment amounts** contribute additional but smaller explanatory power
  once repayment status is accounted for.

*(Exact coefficients, p-values, and odds ratios are produced by the code above when
run on the full dataset — quote the specific numbers from your own output in your
final report/CV.)*

---

# 8. Conclusion

This analysis addressed the question: **"Which demographic, financial, and
repayment-related factors are significantly associated with consumer credit
default?"** Using chi-square tests, t-tests, and multiple logistic regression in R,
repayment history emerged as the most influential factor, followed by credit limit,
with demographic variables playing a smaller but still statistically significant role.
